Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on Swedish and English register data

### 2019-09-25 reports/article.Rmd
Automatic flowchart!

**OBJECTIVE:** Early mortality after total hip arthroplasty (THA) is associated with comorbidity. Diagnosis-based tools such as the Charlson or Elixhauser Comorbidity indices, or the prescription-based RxRisk V score, have been used to quantify this. Those instruments are however complex and requires the use of extensive data sets not routinely available in most doctor-patient settings. We investigated if a simplified model, could substitute those complex indices to predict early mortality after THA.

**PATIENTS AND METHODS:** We studied `r nrow(df)` patients with THA due to primary osteoarthritis `r paste(range(df$P_SurgYear), collapse = " - ")` from the Swedish Hip Arthroplasty Register, linked to national the population register, the National Patient Register and the longitudinal integration database for health insurance and labour market studies from Statistics Sweden. We used stepwise forward and backward logistic regression in combination with bootstrap averaging (bagging), to develop a prediction model for patient deaths within 90 days after surgery. Predictive power was assessed by the area under the reciever operating characteristic curve (AUC). The final model was applied to British data for external validation.

**RESLUTS:** The unadjusted cumulative 90-day survival was `r surv90d`. Best predictive performance for 90-day mortality was found for a model combining `r coefs_print` (AUC = `r with(aucs, AUCtrain_est[name == "full"])`). This mdoel was superior to the established but complex comorbidity measures RxRiskV Score (`r with(aucs, AUCtrain_est[name == "RxRiskV"])`), the Charlson comorbidity index (AUC = `r with(aucs, AUCtrain_est[name == "CCI"])`), and the Elixhauser comorbidity score (AUC = `r with(aucs, AUCtrain_est[name == "ECI"])`).

@@

The presence of pre-surgery comorbidity is associated with poorer outcome after the insertion of total hip arthroplasty (THA). Patients with increased comorbidity an increased risk of early postoperative mortality and revision [@Inacio2015; @Gordon2013], as well as inferior patient-reported outcomes [@Gordon2013; @Hofstede2016]. In research settings, comorbidity is commonly measured using multi-facetted diagnosis- or prescription-based coding algorithms [@Bozic2013]. @Inacio2016 studied the ability of three commonly used coding algorithms to predict mortality after THA and total knee arthroplasty: the Charlson (CCI) and Elixhauser (ECI) Comorbidity Indices, as well as RxRiskV. These comorbidity measures are quite complex to estimate and are based on the availability of extensive datasets including in- and outpatient data on ICD-codes, or detailed information on drug prescriptions prior to surgery. Oftentimes, such datasets can only be created by linking several population-based registries, raising both ethical and practical concerns. Also, each of the comorbidity index exist in numerous versions [@Sundararajan2004; @Deyo1992; @Quan2011; @Cleves1997; @Walraven2009]. Interpretation and comparison between different studies is therefore difficult. 
Comorbidity data have also been used in several universal and arthroplasty-specific risk prediction tools to make risk profiles for individual patients. In the context of trauma, prediction tools are common, and it has been possible to reduce the number of variables without losing predictive power [@Gerdin2016]. No model has so far been broadly accepted for elective THA however [@Manning2016]. An easily applicable tool with few dimensions is thus needed, both in research and in clinical practice. We aimed to find such a model to predict the risk of 90-day postoperative mortality after THA.

>>>>

**OBJECTIVE:** Early mortality after total hip arthroplasty (THA) is associated with comorbidity. Diagnosis-based tools such as the Charlson or Elixhauser Comorbidity indices have been used to quantify this. Those instruments are however complex and requires the use of extensive data sets not routinely available in most doctor-patient settings. We investigated if a simplified model, could substitute those complex indices to predict early mortality after THA.

**PATIENTS AND METHODS:** We studied `r nrow(df)` patients with THA due to primary osteoarthritis `r paste(range(df$P_SurgYear), collapse = " - ")` from the Swedish Hip Arthroplasty Register, linked to the national population register, the National Patient Register and the longitudinal integration database for health insurance and labour market studies from Statistics Sweden. We used a bootstrap ranking procedure using a LASSO-type penalty to develop a prediction model for patient deaths within 90 days after surgery. Predictive power was assessed by the area under the reciever operating characteristic curve (AUC). The final model was applied to British data for external validation.

**RESLUTS:** The unadjusted cumulative 90-day survival was `r surv90d`. Best predictive performance for 90-day mortality was found for a model combining `r coefs_print` (AUC = `r with(aucs, AUCtrain_est[name == "full"])`). This mdoel was superior to the established but complex Charlson comorbidity index (AUC = `r with(aucs, AUCtrain_est[name == "CCI"])`), and the Elixhauser comorbidity score (AUC = `r with(aucs, AUCtrain_est[name == "ECI"])`).

@@

The presence of pre-surgery comorbidity is associated with poorer outcome after the insertion of total hip arthroplasty (THA). Increased comorbidity is associated with an increased risk of early postoperative mortality and revision [@Inacio2015; @Gordon2013], as well as inferior patient-reported outcomes [@Gordon2013; @Hofstede2016]. In research settings, comorbidity is commonly measured using multi-facetted diagnosis- or prescription-based coding algorithms [@Bozic2013]. @Inacio2016 studied the ability of three commonly used coding algorithms to predict mortality after THA and total knee arthroplasty: the Charlson (CCI) and Elixhauser (ECI) Comorbidity Indices, as well as RxRiskV. These comorbidity measures are quite complex to estimate and are based on the availability of extensive datasets including in- and outpatient data on ICD-codes, or detailed information on drug prescriptions prior to surgery. Oftentimes, such datasets can only be created by linking several population-based registries, raising both ethical and practical concerns. Also, each of the comorbidity indices exist in numerous versions [@Sundararajan2004; @Deyo1992; @Quan2011; @Cleves1997; @Walraven2009]. Interpretation and comparison between different studies is therefore difficult. 
Comorbidity data have also been used in several universal and arthroplasty-specific risk prediction tools to make risk profiles for individual patients. In the context of trauma, prediction tools are common, and it has been possible to reduce the number of variables without losing predictive power [@Gerdin2016]. No model has so far been broadly accepted for elective THA however [@Manning2016; @Bulow2017]. An easily applicable tool with few dimensions is thus needed, both in research and in clinical practice. We aimed to find such a model to predict the risk of 90-day postoperative mortality after THA.
